Using numerical modelling to investigate temperature and wave exposure as drivers

### 2019-09-12 literature_review.Rmd
added references to literature review. was still not happy with intro for morphology paper so added more and changed it a fair amount. docx output updated

Dispersal and recruitment are recognised as important processes of marine ecosystems which are driven by a combination of wind and surface currents. This is particulary true for organisms which lack pelagic larvae, and are reliant on other modes of transport to reach new habitats through passive modes of dispersal from anthropogenic and natural sources. Anthropogenic sources include marine litter and structural debris while examples of organic sources include pumice and macroalgae in the form of kelp-rafts. Kelp-rafts have been identified as an important mode of passive dispersal of marine organisms, such as invertebrates [@helmuth1994; whichmann2012], epiphytes [@], grazers [@], invasive species [@] and macroalgae themselves [@] which either make landfall or are transported offshore where they eventually sink. A kelp-raft consists of an entanglement of macroalgae (one or multiple species) which have been dislodged from the benthic environment through hydrodynamic forces (mostly storms) and are positivley buoyant by means of gas-filled pneumatocysts and/or reproductive organs. Kelp-rafts are capable of travelling vast distances [@] and are considered important dispersal vectors in temperate latitudes, such as the Southern California Bight [@hobday2000] and the Northern Baltic Sea [@].

>>>>

Dispersal is recognised as important driver of biodiveristy, composition and structure of ecological systems in the marine environment [@bernardes2018;@helmuth1994;@highsmith1985;macarthur2001;@jackson2010]. These processes are largely dependent on currents for dispersal of migrant populations which ultimately promote connectivity of marine ecosystems [@bernardes2018;@mackas1985;@mackas1985;@zakas2009]. This is particulary true for organisms which lack pelagic larvae, and are reliant on other modes of transport to reach new habitats through passive modes of dispersal from anthropogenic and natural sources. Anthropogenic sources include marine litter and structural debris while examples of organic sources include pumice and macroalgae in the form of kelp-rafts. Kelp-rafts have been identified as an important mode of passive dispersal of marine organisms, such as invertebrates [@helmuth1994; whichmann2012], epiphytes [@], grazers [@], invasive species [@] and macroalgae themselves [@] which either make landfall or are transported offshore where they eventually sink. A kelp-raft consists of an entanglement of macroalgae (one or multiple species) which have been dislodged from the benthic environment through hydrodynamic forces (mostly storms) and are positivley buoyant by means of gas-filled pneumatocysts and/or reproductive organs. Kelp-rafts are capable of travelling vast distances [@] and are considered important dispersal vectors in temperate latitudes, such as the Southern California Bight [@hobday2000] and the Northern Baltic Sea [@].

### 2019-09-12 morphology_paper.Rmd
added references to literature review. was still not happy with intro for morphology paper so added more and changed it a fair amount. docx output updated

Temperature and wave exposure have been shown to play a role in driving kelp morphology. For example, a study by @Serisawa2002-dc compared the morphology of *Ecklonia cava* Kjellman growing in warm temperate and cool temperate morphologies. The results showed that wrinkles in the blade and reduced size to be a characteristic of warm temperate regions. However this study did not take into account the interaction with other environmental variables such as wave exposure. Wave exposure has also been shown to be the main drivers of morphological adaptation across various kelp species. Past research has shown that in highly wave exposed areas kelp morphology tends to take on characteristics which reduce drag, increase strength and increase flexibility [@Denny2002-ob; @Hurd2000; @Fowler-Walker2005, @Thomsen2004-yd]. For example, a study by @Wernberg2005c examined the consistency of wave exposure as a driver of *Ecklonia radiata* (C. Agardh) J. Agardh across a broad geographic range, and  showed trends towards drag-reducing (small size, narrow laterals and blades, low spinosity) and increased strength (large holdfast, thick stipe and thick blades and lamina) at high wave exposure sites. The morphological adaptation to wave exposure must be kept in balance with other important processes, such as nutrient assimilation and light absorption, which are dependent on the amount of surface area of the blades. Therefore, there is a trade-off between reducing overall drag and maintaining nutrient and photosynthetic ability. The morphological adapatibility of kelps is driven by a combination of environmental factors that in turn do not act independently of one another. A study by @Wernberg2003-mp investigated the morphology of *E. radiata* in order to quantify the morphological variation and whether it was dependent on spatial differences along the Australasian coast. They found no correlation between spatial distance and morphological similarity and rather the morphology of kelps was representative of multiple environmental forcings on different morphological characters at different spatial scales [@Wernberg2003-mp]. Due to the complex effects of environmental drivers on kelp morphology, one can also expect differences in morphology between deep and shallow water populations of the same species. Deep and shallow water environments may differ in abiotic processes such as the degree of water coloumn mixing [@Smit2013a], solar heating [@Dellatorre2012; @Dunne2001] and effects of wave dampening [@Kobayashi1993-os; Mia2013; @Dubi1995], all of which do not act idenpendently and instead influence one another.

The kelps *Laminaria pallida* and *Ecklonia maxima* (Osbeck) Papenfuss are important habitat forming seaweeds that exist around the coast of South Africa and offer a unique opportunity to investigate the drivers of macroalgal morphological characteristics between canopy and sub-canopy species, as well as between deep and shallow water environments. Although both these species exist in the subtidal, *L. pallida* dominates deeper waters while *E. maxima* forms dense surface canopies in from the subtidal to shallower waters. Therefore, *E. maxima* is exposed to variations in wind drive surface waves, swell and temperature, while *L. pallida* is exposed to variations in swell and temperature. @Molloy1996-la investigated the effect of depth and wave exposure on *L. pallida* at different depths and wave exposure and showed that depth had a greater effect than wave exposure when considering all the morphological characteristics (wave exposure alters with depth); when considering individual characteristics, however, wave exposure had the most significant effect on blade thickness. Another study by @Rothman2017-pn investigated the changes in morphology in shallow populations of *L. pallida* and *E. maxima* along the South African coastline and into Namibia; *E. maxima* exhibited no morphological changes along the coast but the stipes of *L. pallida* become increasingly hollow further north along the coastline. They suggested that turbidity was the environmental driver responsible for this change.

Measures of wave exposure in ecological studies often incorporate integrative measures of hydrodynamic conditions at a particular site which are based on cartographical models of wave exposure. Cartographical models of wave exposure are regarded as 'fetch-based-models' which measure the length of open water associated with a particular site in a straight line, and are regarded as simple measures of wave exposure. Advances in numerical modelling based on physical/linear wave theory incorporate more complexity (wind forcing, wave-wave interactions, wave breaking, diffraction and variation in wave direction) into the models and allow for a quantitave, reproducible approach for measuring wave exposure. Currently, no ecological studies investigating macroalgal morphological characteristics in relation to the wave environment have incorporated 3D spectral numerical modelling. Previous research investigating drivers of kelp morphological characteristics have often used qualitative estimates of wave exposure and did not consider other morphological characteristics. Furthermore, wind is an important component of surface gravity waves [@Holthuijsen2010-vl] and previous work has not differentiated between surface gravity waves and swell waves, which the current work will aim to achieve. Other considerations such as possible differences in morphology between sub-canopy and canopy species, shallow and deep water populations have not been considered in previous research which ignores the effects of wave damping. The aim of this study is, therefore, to understand how temperature and wave exposure can influence the morphology in two species of kelps around South Africa, as well as between deep and shallow populations of the same species. This will be achieved by initially understanding the variation in temperature and waves using an advanced numerical model. The numerical model will then be used as a basis for investigating the consequences for morphological chracteristics of *E. maxima* and *L. pallida* using multiple statistical tools.

>>>>

Given kelps are cool/temperate organisms which are vulernable to dislodgment under high wave exposure scenarios, changes in ocean temperature and an increase in storm frequency and magnitude pose the biggest direct threats to their survival. Warmer temperatures reduces the resilience of kelp individuals to perbutations [@wernberg2010], causes fragmentation through weakening tissues [@simonson2015] and reduces growth and productivity [@bearham2013; @gerard1997; @gao2013; @zimmerman1986]; while storms and high wave energy dislodge and break kelps [@byrnes2011; @graham2004; @seymour1989]. Despite these threats and disturbances kelps persist which across a broad range of environments which is largely due to their morphological plasticity [@fowler2006; @reed2003]. A study by @Wernberg2003-mp investigated the morphology of *E. radiata* in order to quantify the morphological variation and whether it was dependent on spatial differences along the Australasian coast. They found no correlation between spatial distance and morphological similarity and rather the morphology of kelps was representative of multiple environmental forcings on different morphological characters at different spatial scales [@Wernberg2003-mp].Due to the complex effects of environmental drivers on kelp morphology, one can also expect differences in morphology between deep and shallow water populations of the same species. Deep and shallow water environments may differ in abiotic processes such as the degree of water column mixing [@Smit2013a], solar heating [@Dellatorre2012; @Dunne2001] and effects of wave dampening [@Kobayashi1993-os; Mia2013; @Dubi1995].

Although other factors play a role in kelp morphological adaptation, wave exposure and temperature have been identified as the main drivers of morphological adaptation across various kelp species [@bekkby2014; @wernberg2005; @Denny2002-ob; @Hurd2000; @Fowler-Walker2005; @Thomsen2004-yd; @mabin2013]. For example, a study by @Serisawa2002-dc compared the morphology of *Ecklonia cava* Kjellman growing at warmer sites and cooler sites, and showed wrinkling of the blade and reduced size to be a characteristic of higher temperature ranges. Temperature affects important kelp physiological processes such as photosynthesis and respiration, which ultimately influences growth and productivity resulting in reduced size [@bearham2013; @gerard1997; @gao2013; @zimmerman1986]. Past research has shown that in highly wave exposed areas kelp morphology tends to take on characteristics which reduce drag, increase strength and increase flexibility [@Denny2002-ob; @Hurd2000; @Fowler-Walker2005; @Thomsen2004-yd]. For example, a study by @Wernberg2005c examined the consistency of wave exposure as a driver of *Ecklonia radiata*  (C. Agardh) J. Agardh. across a broad geographic range, and showed trends towards drag-reducing (small size, narrow laterals and blades, low spinosity) and increased strength (large holdfast, thick stipe and thick blades and lamina) at high wave exposure sites.

The kelps *Laminaria pallida* and *Ecklonia maxima* (Osbeck) Papenfuss are important habitat forming seaweeds that exist around the coast of South Africa and offer a unique opportunity to investigate the drivers of macroalgal morphological characteristics between canopy and sub-canopy species, as well as between deep and shallow water environments. Although both these species exist in the subtidal, *L. pallida* dominates deeper waters while *E. maxima* forms dense surface canopies in the subtidal to shallower waters. Therefore, *E. maxima* is exposed to variations in wind driven surface waves, swell and temperature, while *L. pallida* is exposed to variations in swell and temperature. @Molloy1996-la investigated the effect of depth and wave exposure on *L. pallida* at different depths and wave exposure and showed that depth had a greater effect than wave exposure when considering all the morphological characteristics (wave exposure alters with depth); when considering individual characteristics, however, wave exposure had the most significant effect on blade thickness. Another study by @Rothman2017-pn investigated the changes in morphology in shallow populations of *L. pallida* and *E. maxima* along the South African coastline and into Namibia; *E. maxima* exhibited no morphological changes along the coast but the stipes of *L. pallida* become increasingly hollow further north along the coastline. They suggested that turbidity in relation to light attenuation was the environmental driver responsible for this change.

Measures of wave exposure in ecological studies often incorporate integrative measures of hydrodynamic conditions at a particular site which are based on cartographical models of exposure. Cartographical models of wave exposure are 'fetch-based-models' which measure the length of open water associated with a particular site in a straight line, and are regarded as simple measures of wave exposure. Advances in numerical modelling based on physical/linear wave theory incorporate more complexity (wind forcing, wave-wave interactions, wave breaking, diffraction and variation in wave direction) into the models and allow for a quantitave, reproducible approach for quantifying wave exposure. Although @mabin2013 investigated kelp morphological characteristics using a 2 dimensional WAM model, currently no studies investigating macroalgal morphological characteristics in relation to the wave environment have incorporated 3D spectral numerical modelling. Furthermore no current studies have identified specific wave metrics in driving kelp morphological charactersitcs using advanced numerical model, which the current work will aim to achieve. Other considerations such as possible differences in morphology between sub-canopy and canopy species, shallow and deep water populations have not been considered using advanced numerical and statisical tools. The aim of this study is, therefore, to understand how temperature and wave exposure can influence the morphology in two species of kelps around South Africa, as well as between deep and shallow populations of the same species. This will be achieved by initially understanding the variation in temperature and waves using an advanced numerical model. The numerical model will then be used as a basis for investigating the consequences for morphological chracteristics of *E. maxima* and *L. pallida* using advanced statistical tools.
